### 2020-06-18 report/sections/server/extensions/error.tex
Report: final proofread

The WebSocket \texttt{close()} method accepts a 
\textit{close code}, which is analogue to exit codes for
executables. We extend this mechanism to distinguish
between the different variants of session cancellation,
as motivated in \cref{subsection:errormotivation}.
As shown in \cref{fig:cancelenum},
we define unique close codes for different types
of session cancellation under \filename{Cancellation.ts}
for both server-side (\cref{subfig:cancelenumsvr})
and browser-side endpoints (\cref{subfig:cancelenumclient}) 
respectively. 
To respect the convention specified in the WebSocket protocol
RFC \cite{WebSocketRFC} that the 1xxx close codes 
are reserved for internal use,
we define our custom close codes from 4000 onwards.
The enum values are named in a self-explanatory way.

>>>>

The WebSocket \texttt{close()} method accepts a 
\textit{close code}, which is analogue to exit codes for
executables. We extend this mechanism to distinguish
between the different variants of session cancellation,
as motivated in \cref{subsection:errormotivation}.
As shown in \cref{fig:cancelenum},
we define unique close codes for different types
of session cancellation under \filename{Cancellation.ts}
for both server-side (\cref{subfig:cancelenumsvr})
and browser-side endpoints (\cref{subfig:cancelenumclient}) 
respectively. 
To respect the convention specified in the WebSocket Protocol
\cite{WebSocketRFC} that the 1xxx close codes 
are reserved for internal use,
we define our custom close codes from 4000 onwards.
The enum values are named in a self-explanatory way.

### 2020-06-18 report/sections/server/node.tex
Report: final proofread

We discuss the challenges of implementing
session types on the Node.js runtime 
(\cref{section:nodechallenges})
and motivate our approach (\cref{section:nodeapproach}).
We explain our design choices for encoding
the EFSM for server-side endpoints (\cref{section:nodeefsm}),
and detail the session runtime that is designed
to execute the EFSM on Node.js (\cref{section:noderuntime}).
We conclude by analysing some alternative designs 
(\cref{section:nodealt}) and evaluating the limitations 
of our approach (\cref{section:nodelimitations}).

\begin{figure}[!b]
\centering
\includegraphics[width=0.6\textwidth]{AdderSvrFSM}
\captionof{figure}{\trole{Svr} Endpoint FSM
in \tprotocol{Adder} protocol}
\label{fig:addersvrfsm}
\end{figure}

>>>>

We discuss the challenges of implementing
session types on the Node.js runtime 
(\cref{section:nodechallenges})
and motivate our approach (\cref{section:nodeapproach}).
We explain our design choices for encoding
the EFSM for server-side endpoints (\cref{section:nodeefsm}),
and present a session runtime designed
to execute the EFSM on Node.js (\cref{section:noderuntime}).
We conclude by analysing some alternative designs 
(\cref{section:nodealt}) and evaluating the limitations 
of our approach (\cref{section:nodelimitations}).

\begin{figure}[!b]
\centering
\includegraphics[width=0.6\textwidth]{AdderSvrFSM}
\captionof{figure}{\trole{Svr} Endpoint FSM
in \tprotocol{Adder} Protocol}
\label{fig:addersvrfsm}
\end{figure}

### 2020-06-18 report/sections/server/node/efsm.tex
Report: final proofread

Handler APIs act as the \textit{seams} of the EFSM.
A seam is ``a place where one can alter the
behaviour in the program without editing in that place''
\cite{seam}. The developer implements handlers
to define what message to send and how to process a received
message, without access to the send and receive APIs
that are used to execute the EFSM.
We collect these APIs under the \texttt{Handler} namespace. 
We introduce the generated handlers for sending and receiving states.
These are non-terminal states that will involve the encoding of its
successor state. The reader will notice that, in the listings below,
the successor state is stated to be under the
\texttt{Implementation} namespace: we explain in 
\cref{subsection:nodeefsmimplementation}, but for now,
it is sufficient to acknowledge that those refer to the encoding
of the successor state.
As a design choice, we \textit{do not} generate handlers for
terminal states, because the semantics of inactivity mean
there is nothing to handle.

@@

We attempt to address this by defining an enum of state identifiers
for each type of state (i.e. an enum for send states, 
an enum for receive states)
upon which to execute the EFSM. 
This can solve the switch statement problem.
Now, we are left with defining a mapping between the 
state identifier enum to the handler type. This construct would be
analogue to \textit{dependent types}, which again, is not a feature
of the TypeScript type system.

@@

Unfortunately, this approach does not work for the simple fact that
conditional types were not designed to be exploited in this manner.
The main limitation of conditional types is its 
\textit{distributivity} when the type ``parameter'' is an
union type (which is the case for enums, as
\texttt{S = SendState.S53 | SendState.S54| ...}), where

@@

We discuss our runtime implementation shortly 
(\cref{section:noderuntime}), where we disclose more details regarding
the role of the \texttt{Implemenation} wrapper API in
the runtime.

>>>>

Handler APIs act as the \textit{seams} of the EFSM.
A seam is ``a place where one can alter the
behaviour in the program without editing in that place''
\cite{seam}. The developer implements handlers
to define what message to send and how to process a received
message, without accessing the send and receive APIs
that are used to execute the EFSM.
We collect these APIs under the \texttt{Handler} namespace. 
We introduce the generated handlers for sending and receiving states.
These are non-terminal states that will involve the encoding of its
successor state. The reader will notice that, in the listings below,
the successor state is stated to be under the
\texttt{Implementation} namespace: we explain in 
\cref{subsection:nodeefsmimplementation}, but for now,
it is sufficient to acknowledge that those refer to the encoding
of the successor state.
As a design choice, we \textit{do not} generate handlers for
terminal states, because the semantics of inactivity mean
there is nothing to handle.

@@

We \textit{attempt} to address this by defining an enum of state identifiers
for each type of state (i.e. an enum for send states, 
an enum for receive states)
upon which to execute the EFSM. 
This can solve the switch statement problem.
Now, we are left with defining a mapping between the 
state identifier enum to the handler type. This construct would be
analogue to \textit{dependent types}, which again, is not a feature
of the TypeScript type system.

@@

Unfortunately, this approach \textbf{does not work}
for the simple fact that
conditional types were not designed to be exploited in this manner.
The main limitation of conditional types is its 
\textit{distributivity} when the type parameter is an
union type (which is the case for enums, as
\texttt{S = SendState.S53 | SendState.S54 | ...}), where

@@

We discuss our runtime implementation shortly 
(\cref{section:noderuntime}), where we disclose more details regarding
the role of the \texttt{Implementation} wrapper API in
the runtime.

### 2020-06-18 report/sections/server/node/limitations.tex
Report: final proofread

Our APIs also rely on state identifiers, which also
does not help code readability, as it is neither apparent nor
self-documenting what \texttt{S51} refers to -- some existing work
\cite{Hybrid2016} augment the state identifiers with the channel
action and labels involved (e.g. 
\texttt{S51_Receive__Add_Number__Quit_String}).

>>>>

Our APIs also rely on state identifiers, which also
does not help code readability, as it is neither apparent nor
self-documenting what \texttt{S51} refers to -- the work of
Hu and Yoshida
\cite{Hybrid2016} augment
the state identifiers with the channel
action and labels involved (e.g. 
\texttt{S51_Receive__Add_Number__Quit_String}).

### 2020-06-18 report/sections/server/node/runtime.tex
Report: final proofread

We define the session runtime for the \trole{Svr} endpoint
of the \tprotocol{Adder} protocol in \filename{Svr.t}s,
named after the endpoint.
It exposes a \textbf{public API}
with seams for the developer to pass in the WebSocket server 
and application logic
(i.e. their handler implementations).
It is developer's responsibility to construct the
WebSocket server and set it up to listen for incoming connections.
Internally, it keeps a \textit{private API}
for executing the EFSM, when all participants have 
joined the session.

@@

	// Ignore if role is aleady taken
	if (!waiting.has(role)) { return socket.close(); } (*@\label{line:occupied}@*)

@@

We get the label, payload and successor implementation
directly from the handler implemented by the developer, 
accurately typed by how we define the handler API in \filename{EFSM.ts}.
The developer does not need to specify which role
to send to: this is a sensible design choice, as we know this
from the Scribble protocol, so we do not need the developer 
to specify separately.
As a result, we generate the code to send the message
to the correct role (\cref{line:nodesendrole}).
We use the \texttt{send()} method passed down
by the runtime to commit our communication action:
the runtime will handle how to serialise the message and perform
the send. We guarantee that \texttt{send()} is called
\textbf{exactly once} by construction, thus channel linearity
is never violated.
Finally, we use the parameterised EFSM transition handler
to notify the rutime which specific state to transition to.

@@

They are decoded using the same interface schema on the receiving end
using \texttt{JSON.parse()}: whilst the method return type is the
dynamic \lstonelinejs{any} type, 
we guarantee type safety by construction
as we performed the serialisation in the first place, so
we can safely interpret the deserialised content 
using a concrete type.

@@

class Session {
	// Optional type, same as `MessageHandler | undefined`
	private handler?: MessageHandler; | undefined(*@\label{line:noderuntimehandler}@*)

@@

@@

We observe that message arrivals do not have to be causally related.
However, if we consider a similar \textit{binary} example
\[
A \to S: \text{\lstonelinejs{M1(string)}}. 
A \to S: \text{\lstonelinejs{M2(number)}}. \texttt{end}
\]
then \tmsg{M1} \textit{must} arrive before \textit{M2},
since this is sent through the WebSocket connection between 
\trole{A} and \trole{S}, and FIFO guarantees are respected 
for each individual WebSocket connection. 
We visualise the possible orders of message receive events
in \cref{fig:nodereceivecompare}.

\begin{figure}[!h]
\centering
\begin{subfigure}[b]{0.8\textwidth}
\centering
\includegraphics[width=\textwidth]{NodeRuntimeReceive2}
\caption{Message processed before transitioning to receive state}
\label{subfig:nodereceivemsgfirst}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.8\textwidth}
\centering
\includegraphics[width=\textwidth]{NodeRuntimeReceive1}
\caption{Message processed after transitioning to receive state}
\label{subfig:nodereceivehandlefirst}
\end{subfigure}
\caption{Possible Orderings for Handling Message Event and Preparing
Receive State}
\label{fig:nodereceivecompare}
\end{figure}

>>>>

We define the session runtime for the \trole{Svr} endpoint
of the \tprotocol{Adder} protocol in \filename{Svr.t}s,
named after the endpoint.
It exposes a \textbf{public API}
with seams for the developer to pass in the WebSocket server 
and application logic
(i.e. implementations of the handler APIs).
It is developer's responsibility to construct the
WebSocket server and set it up to listen for incoming connections.
Internally, it keeps a \textit{private API}
for executing the EFSM, when all participants have 
joined the session.

@@

	// Ignore if role is already taken
	if (!waiting.has(role)) { return socket.close(); } (*@\label{line:occupied}@*)

@@

We get the label, payload and successor implementation
directly from the handler implemented by the developer, 
accurately typed by how we define the handler API in \filename{EFSM.ts}.
The developer does not need to specify which role
to send to: this is a sensible design choice, as we know this
from the Scribble protocol, so we do not need the developer 
to specify separately.
As a result, we generate the code to send the message
to the correct role (\cref{line:nodesendrole}).
We use the \texttt{send()} method passed down
by the runtime to commit our communication action:
the runtime will handle how to serialise the message and perform
the send. We guarantee that \texttt{send()} is called
\textbf{exactly once} by construction, thus channel linearity
is never violated.
Finally, we use the parameterised EFSM transition handler
to notify the runtime which specific state to transition to.

@@

They are decoded using the same interface schema on the receiving end
using \texttt{JSON.parse()}.
Whilst the method return type is the
dynamic \lstonelinejs{any} type, 
we guarantee type safety by construction
as we performed the serialisation in the first place, so
we can safely interpret the deserialised content 
using a concrete type.

@@

class Session {
	// Optional type, same as `MessageHandler | undefined`
	private handler?: MessageHandler; (*@\label{line:noderuntimehandler}@*)

@@

\[
A \to S: \text{\lstonelinejs{M1(string)}}.~ 
B \to S: \text{\lstonelinejs{M2(number)}}.~\texttt{end} 
\]

@@

We observe that message arrivals do not have to be causally related.
However, if we consider a similar \textit{binary} example
\[
A \to S: \text{\lstonelinejs{M1(string)}}.~ 
A \to S: \text{\lstonelinejs{M2(number)}}.~\texttt{end}
\]
then \tmsg{M1} \textit{must} arrive before \textit{M2},
since this is sent through the WebSocket connection between 
\trole{A} and \trole{S}, and FIFO guarantees are respected 
for each individual WebSocket connection. 
We visualise the possible orders of message receive events
in \cref{fig:nodereceivecompare}.

\begin{figure}[!h]
\centering
\begin{subfigure}[b]{0.8\textwidth}
\centering
\includegraphics[width=\textwidth]{NodeRuntimeReceive2}
\caption{Message processed before transitioning to receive state}
\label{subfig:nodereceivemsgfirst}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.8\textwidth}
\centering
\includegraphics[width=\textwidth]{NodeRuntimeReceive1}
\caption{Message processed after transitioning to receive state}
\label{subfig:nodereceivehandlefirst}
\end{subfigure}
\caption{Possible Orderings for Receiving Message and
Registering Handler}
\label{fig:nodereceivecompare}
\end{figure}

### 2020-06-18 report/sections/server/react.tex
Report: final proofread

We highlight the challenges of session-typed GUI programming
(\cref{section:reactchallenges}) and motivate 
our approach (\cref{section:reactapproach})
for adapting the proposal from
\cite{MVU2020} for the React.js framework, along with a brief
introduction to the framework.
We proceed to explain how we leverage React
to encode the EFSM (\cref{section:reactefsm})
and implement the session runtime (\cref{section:reactruntime}),
focusing on our novel approach to preventing
channel reuse in GUI programming using React.
We conclude by analysing some alternative designs
(\cref{section:reactalt})
and evaluating the limitations of our approach
(\cref{section:reactlimitations}).

\begin{figure}[!b]
\centering
\includegraphics[width=0.6\textwidth]{AdderClientFSM}
\captionof{figure}{\trole{Client} Endpoint FSM
in \tprotocol{Adder} protocol}
\label{fig:adderclientfsm}
\end{figure}

>>>>

We highlight the challenges of session-typed GUI programming
(\cref{section:reactchallenges}) and motivate 
our approach (\cref{section:reactapproach})
for adapting the proposal from
\cite{MVU2020} for the React.js framework, along with a brief
introduction to the framework.
We proceed to explain how we leverage React
to encode the EFSM (\cref{section:reactefsm})
and implement the session runtime (\cref{section:reactruntime}),
focusing on our novel approach for preventing
channel reuse in GUI programming using React.
We conclude by analysing alternative designs
(\cref{section:reactalt})
and evaluating the limitations of our approach
(\cref{section:reactlimitations}).

\begin{figure}[!b]
\centering
\includegraphics[width=0.6\textwidth]{AdderClientFSM}
\captionof{figure}{\trole{Client} Endpoint FSM
in \tprotocol{Adder} Protocol}
\label{fig:adderclientfsm}
\end{figure}

### 2020-06-18 report/sections/server/react/approach.tex
Report: final proofread

We implement model types for the EFSM on top of the 
\emph{React.js} (React) framework developed by Facebook \cite{React}.
React is widely used in industry to create scalable single-page
web applications, so this makes our workflow beneficial in an
industrial context. 
The framework defines a way for data to flow
between UI elements, and empowers the UI to subscribe and
``react'' to data changes.
We aim to implement similar behaviour with respect to the EFSM:
the UI should react to EFSM state transitions,
so we can \textbf{statically} ensure that the
channel actions ``present'' on the browser at any given time
are those permitted by the current EFSM state.
We introduce the framework in \cref{subsection:react}.

When executing \fancyname{SessionTS} to generate code
for the \trole{Client} endpoint specifying the
\texttt{browser} target, the developer obtains the following 
groups of files:

>>>>

We implement model types for the EFSM on top of the 
\emph{React.js} (React) framework developed by Facebook \cite{React}.
React is widely used in industry to create scalable single-page
web applications, so this makes our workflow beneficial in an
industrial context. 
The framework defines a way for data to flow
between UI elements, and empowers the UI to subscribe and
``react'' to data changes;
we introduce the framework in \cref{subsection:react}.
We aim to implement similar behaviour with respect to the EFSM:
the UI should react to EFSM state transitions,
so we can \textbf{statically} ensure that the
channel actions ``present'' on the browser at any given time
are those permitted by the current EFSM state.

When executing \reactcodegen to generate code
for the \trole{Client} endpoint specifying the
\texttt{browser} target, the developer obtains the following 
groups of files:

### 2020-06-18 report/sections/server/react/challenges.tex
Report: final proofread

Our goal with \fancyname{ReactMPST} is to
generate session type implementations for the web browser.
Integrating session types into user interfaces is inherently difficult --
assuming channel actions are bound to user interface events 
(i.e. clicking a button sends a message), how does one formalise
channel linearity? How do we guarantee that, if the button
triggers a send at some EFSM state, that it triggers not more than
one channel action, \textit{and} the user cannot trigger the action
at another EFSM state?

We recap how existing work 
(as discussed in \cref{subsection:sessiontypewebdev})
tackle browser-side session typing.
Fowler introduced the concept of model types to prevent
channel linearity violation in his proposal for
integrating session types with
graphical user interface (GUI) programming \cite{MVU2020}, but
it uses the Links web programming language \cite{LINKS} which
lacks compatibility with the ecosystem of JavaScript
libraries that one might use in front-end development as well.
The session type-safe web development framework presented
by King et al. \cite{PureScript2019} generates APIs for
a functional target language in PureScript, and relies on
the \textit{Concur UI} framework that constructs UIs sequentially.

As motivated in \cref{section:intro},
we find these proposals to come at the cost of
limiting developer productivity by adopting unconventional practices
that may require a learning curve.
Through our work, we aim to distil the key
concepts from existing work that provide session type safety
for web-based GUI programming, and implement them using
mainstream front-end web development tools (i.e. TypeScript
and React), to provide developers with an intuitive way to implement
browser-side endpoints that guarantee communication safety.

>>>>

Our goal with \reactcodegen is to
generate session type implementations for the web browser.
Integrating session types into user interfaces is inherently difficult --
assuming channel actions are bound to user interface events 
(i.e. clicking a button sends a message), how does one formalise
channel linearity? How do we guarantee that, if the button
triggers a send at some EFSM state, that it triggers not more than
one channel action, \textit{and} the user cannot trigger the action
at another EFSM state?

We recap how existing work 
(as discussed in \cref{subsection:sessiontypewebdev})
tackle browser-side session typing.
Fowler \cite{MVU2020} introduced the concept of model types to 
prevent
channel linearity violation in his proposal for
integrating session types with
GUI programming, but
it uses the Links web programming language \cite{LINKS} which
lacks compatibility with the ecosystem of JavaScript
libraries that one might use in front-end development as well.
The session type-safe web development framework presented
by King et al. \cite{PureScript2019} generates APIs for
a functional target language in PureScript, and relies on
the \textit{Concur UI} framework that constructs UIs sequentially.

As motivated in \cref{section:intro},
we find these proposals to come at the cost of
limiting developer productivity by adopting unconventional practices
that may require a learning curve.
Through our work, we aim to distil the key
concepts from \cite{MVU2020,LINKS,PureScript2019} 
that provide session type safety
for web-based GUI programming, and implement them using
mainstream front-end web development tools (specifically TypeScript
and React), to provide developers with an intuitive way to implement
browser-side endpoints that guarantee communication safety.

### 2020-06-18 report/sections/server/react/efsm.tex
Report: final proofread

We show the developer's implementation 
of a send state for the \trole{Client} endpoint
from the \tprotocol{Adder} protocol in \cref{lst:adderclientsenddev}.
Component state is an implementation detail 
that is not related to EFSM execution, so we allow
the developer to customise this through specifying an 
interface (\cref{line:clientsendstatestate}).

@@

The factory API for binding send actions to UI elements appear on
\cref{line:clientsendfactory1,line:clientsendfactory2}.
For example, \cref{line:clientsendfactory1} reads: 
\textit{build a React component that sends the \tmsg{ADD}
message with \texttt{[num1, num2]} as payload,
when the user clicks on it}.
\tmsg{ADD} is the factory API we generate: it takes an event identifier
and an event handler which must return the payload 
-- \textit{appropriately typed with respect to the Scribble protocol} -- 
for sending the \tmsg{ADD} message.
Swapping \cref{line:clientsendpayload1,line:clientsendpayload2}
will result in a compile-time type error due to invalid 
payload types, which is the intended behaviour.

@@

Contrary to our discussions about modelling dependent types in
\cref{subsection:nodeefsmimplementation}, we can do this here
because: \textbf{(1)} we are using index type queries instead
of conditional types, and \textbf{(2)} we know the type
of \texttt{ev} at compile-time from the developer's implementation,
so we are not resolving index type queries with union types.

@@

		this.QUIT = props.factory<[string]>( (*@\label{line:factory2}@*)
			'QUIT', ReceiveState.S43);
	}
}
\end{lstlisting}
\captionof{lstlisting}{Generated Type for \trole{Client} Send State
in \tprotocol{Adder} protocol}
\label{lst:reactsendcomponent}
\end{figure}

@@

	handle(message: any): State { (*@\label{line:reactreceivehandle}@*)
		const parsedMessage = JSON.parse(message) as Message;
		switch (parsedMessage.label) { (*@\label{line:reactreceivediscriminate}@*)
			case Labels.THANKS:
				this.THANKS(...parsedMessage.payload);
				return TerminalState.S41;
			case Labels.TERMINATE:
				this.TERMINATE(...parsedMessage.payload);
				return TerminalState.S41;
		}
	}
}
\end{lstlisting}
\captionof{lstlisting}{Generated Code for \trole{Client} 
Receive State S42 in \tprotocol{Adder} protocol}
\label{lst:reactreceivecomponent}
\end{figure}

>>>>

We show the developer's implementation 
of a send state for the \trole{Client} endpoint
from the \tprotocol{Adder} protocol in \cref{lst:adderclientsenddev}.
Component state is an implementation detail 
that is not related to EFSM execution, so we allow
the developer to customise this through specifying an 
interface (\cref{line:clientsendstatestate}) as the generic type.

@@

The factory API for binding send actions to UI elements appear on
\cref{line:clientsendfactory1,line:clientsendfactory2}.
For example, \cref{line:clientsendfactory1} reads: 
\textit{build a React component that sends the \tmsg{ADD}
message with \texttt{[num1, num2]} as payload,
when the user clicks on it}.
\tmsg{ADD} is the factory API we generate: it takes an event identifier
and an event handler which must return the payload 
-- \textit{appropriately typed with respect to the Scribble protocol} -- 
for sending the \tmsg{ADD} message.
Swapping \cref{line:clientsendpayload1,line:clientsendpayload2}
will result in a compile time type error due to invalid 
payload types, which is the intended behaviour.

@@

Contrary to our discussions about modelling dependent types in
\cref{subsection:nodeefsmimplementation}, we can do this here
because: \textbf{(1)} we are using index type queries instead
of conditional types, and \textbf{(2)} we know the type
of \texttt{ev} at compile time from the developer's implementation,
so we are not resolving index type queries with union types.

@@

		this.QUIT = props.factory<[string]>( (*@\label{line:factory2}@*)
			'QUIT', ReceiveState.S43);
	}
}
\end{lstlisting}
\captionof{lstlisting}{Generated Type for \trole{Client} Send State
in \tprotocol{Adder} Protocol}
\label{lst:reactsendcomponent}
\end{figure}

@@

	handle(message: any): State { (*@\label{line:reactreceivehandle}@*)
		const parsedMessage = JSON.parse(message) as Message;
		switch (parsedMessage.label) { (*@\label{line:reactreceivediscriminate}@*)
			case Labels.THANKS:
				this.THANKS(...parsedMessage.payload);
				return TerminalState.S41;
			case Labels.TERMINATE:
				this.TERMINATE(...parsedMessage.payload);
				return TerminalState.S41;
		}
	}
}
\end{lstlisting}
\captionof{lstlisting}{Generated Code for \trole{Client} 
Receive State S42 in \tprotocol{Adder} Protocol}
\label{lst:reactreceivecomponent}
\end{figure}

### 2020-06-18 report/sections/server/react/runtime.tex
Report: final proofread

There are many parallels between the runtime we generate
for \fancyname{ReactMPST} compared to that of \fancyname{NodeMPST}.
We focus on the details unique to React endpoints.

@@

\begin{lstlisting}[language=javascript]
type Props = {
	endpoint: string,
	states: {						// mapping of state identifier
		S40: Constructor<S40>,		// to EFSM state components
		S42: Constructor<S42>,		// implemented by developer
		S43: Constructor<S43>,
		S41: Constructor<S41>,
	},
	waiting: React.ReactNode,		// render when waiting to start
	connectFailed: React.ReactNode,	// render if connection failed
};

export class Session extends React.Component<Props, ...
\end{lstlisting}

The developer provides the WebSocket URL for the runtime
to instantiate the connection.
As for the EFSM state implementations derived from the abstract
classes, the developer constructs an object to map state identifiers
to concrete implementations. 
The developer only needs to pass the \textit{constructor} function,
as the runtime will instantiate the React components separately
and pass the required props based on the type of state.
Additionally, the developer needs to define what to render
whilst \texttt{waiting} for the session to begin, and possibly
an error screen if the connection has failed.
We show an example below.

@@

// Main application component
class MainApp extends React.Component {
	render() {
		return <div>
			<h1>Adder Client</h1>
			<Client
				endpoint='ws://localhost:8080'
				states={{
					S40: InputWindow,
					S41: QuitReceived,
					S42: WaitScreen,
					S43: PendingQuit,		
				}}>
				waiting={<h1>Pending Connection</h1>}
				connectFailed={<p>Connection Failed</p>}
			/>
		</div>;
	}
}
\end{lstlisting}

@@

We take the same approach as \fancyname{NodeMPST} --
the runtime keeps track of the ``active'' receive handler,
and the \texttt{onmessage} event listener for the WebSocket
is defined to dynamically process the incoming message
using the active receive handler.
We also need to address the same consistency problem, since
a message arrival event can be handled \textit{before}
the receive handler is registered.
The receive handler is registered when the receive state component
is rendered on the DOM, which occurs when \lstonelinejs{this.setState}
is invoked in the transition function.
\lstonelinejs{this.setState} may be \textit{asynchronous}
as React will internally batch multiple calls into a single update
for performance \cite{ReactState}, so it is possible for the WebSocket message
event to be queued (and hence, procesesd) before the receive handler
is registered.

@@

private receive({ data }: MessageEvent) {
	const handler = this.handlerQueue.shift();
	if (handler !== undefined) { this.advance(handler(data); } (*@\label{line:reactreceiveadvance2}@*)
	else { this.messageQueue.push(data); }
}
\end{lstlisting}
\captionof{lstlisting}{Receive Handler Registration and Message Event
Listener in \fancyname{ReactMPST}}
\label{lst:reactreceiveruntime}
\end{figure}

>>>>

There are many parallels between the runtime we generate
for \reactcodegen compared to that of \nodecodegen.
We focus on the details unique to React endpoints.

@@

\begin{lstlisting}[language=javascript]
type Props = {
	endpoint: string,				// WebSocket URL to connect to
	states: {						// mapping of state identifier
		S40: Constructor<S40>,		// to EFSM state components
		S42: Constructor<S42>,		// implemented by developer
		S43: Constructor<S43>,
		S41: Constructor<S41>,
	},
	waiting: React.ReactNode,		// render when waiting to start
	connectFailed: React.ReactNode,	// render if connection failed
};

export class Session extends React.Component<Props> { ... }
\end{lstlisting}

The developer provides the WebSocket URL for the runtime
to instantiate the connection.
As for the EFSM state implementations derived from the abstract
classes, the developer constructs an object to map state identifiers
to concrete implementations. 
The developer only needs to pass the \textit{constructor function},
as the runtime will instantiate the React components separately
and pass the required props based on the type of state.
Additionally, the developer needs to define what to render
whilst \texttt{waiting} for the session to begin, and possibly
an error screen if the connection has failed.
We show an example below.

@@

// Main application component
class MainApp extends React.Component {
	render() {
		return <div>
			<h1>Adder Client</h1>
			<Session
				endpoint='ws://localhost:8080'
				states={{
					S40: InputWindow,
					S41: QuitReceived,
					S42: WaitScreen,
					S43: PendingQuit,		
				}}>
				waiting={<h1>Pending Connection</h1>}
				connectFailed={<p>Connection Failed</p>}
			/>
		</div>;
	}
}
\end{lstlisting}

@@

We take the same approach as \fancyname{NodeMPST} --
the runtime keeps track of the ``active'' receive handler,
and the \texttt{onmessage} event listener for the WebSocket
is defined to dynamically process the incoming message
using the active receive handler.
We also need to address the same consistency problem, since
a message arrival event can be handled \textit{before}
the receive handler is registered.
The receive handler is registered when the receive state component
is rendered on the DOM, which occurs when \lstonelinejs{this.setState}
is invoked in the transition function.
\lstonelinejs{this.setState} may be \textit{asynchronous}
as React will internally batch multiple calls into a single update
for performance \cite{ReactState}, so it is possible for the WebSocket message
event to be queued (and hence, processed) before the receive handler
is registered.

@@

private receive({ data }: MessageEvent) {
	const handler = this.handlerQueue.shift();
	if (handler !== undefined) { this.advance(handler(data); } (*@\label{line:reactreceiveadvance2}@*)
	else { this.messageQueue.push(data); }
}
\end{lstlisting}
\captionof{lstlisting}{Receive Handler Registration and Message Handler
in \fancyname{ReactMPST}}
\label{lst:reactreceiveruntime}
\end{figure}
